### 2020-09-07 tex/design.tex
first commit

As described in Section \ref{sec:related}, FrodoKEM requires heavy use of PRNG / seed expanding. In the algorithm specifications, it is suggested to use either SHAKE or AES. In particular, the most computationally intensive operations, Line 7 of Algorithm \ref{alg:encaps}, requires 410k or 953k 16-bit pseudo-random values, depending on the parameter set used. In order for PRNG not to be the bottle-neck it needs to achieve a very high throughput (ideally with relatively low area consumption) typically in the range of 16 bits per clock cycle. In a previous hardware design, proposed by Howe et al. \cite{howe2018standard}, high throughput for the PRNG was achieved by pre-calculating randomness and storing it in BRAM. Random data newly calculated was then written into the memory, overwriting the random data previously stored. This is an efficient approach, however a more efficient PRNG that would not require BRAM usage, would have the potential to increase the operating frequency of the design and thus improve its throughput. Moreover, parallelisations was not possible for this design, as this would require a faster SHAKE design, increasing the area consumption by 3-8x \cite{bertoni2012keccak}. The area consumption of SHAKE (or AES) was an issue with the previous design. For example, cSHAKE used within FrodoKEM-640 Encaps occupies 42\% of the overall hardware resources~\cite{howe2018standard}.

@@

In order to fully explore the potential of FrodoKEM in hardware, we propose several architectures characterized by different design goals (in terms of throughput). We use the proposed architecture to implement key generation, encapsulation, and decapsulation, on two sets of parameters proposed in the specifications: FrodoKEM-640 and FrodoKEM-976. Our designs use 1x, 4x, 8x, and 16x parallel multiplications during the most computationally intensive parts in FrodoKEM. These operations are the LWE matrix multiplications of the form:
%The following is the LWE calculation of the type: 
\begin{equation}
\mathbf{B} = \mathbf{S} \mathbf{A} + \mathbf{E}, \label{lwe}
\end{equation}
 required in key generation, encapsulation, and decapsulation. In the previous hardware implementations of FrodoKEM, the operations of the type of Equation \ref{lwe} took approximately 97.5\% of the overall run-time of the designs \cite{howe2018standard}. As in the literature, we exploit DSP slices on the FPGA for the multiply-and-accumulate (MAC) operations required for matrix multiplication. Hence, each parallel multiplication of the proposed designs requires its own DSP slice.

\begin{figure}[htbp]\centering
    %\advance\leftskip-1cm
\includegraphics[scale=1]{figures/arch_new.pdf}
\caption{A high-level overview of the proposed hardware designs for FrodoKEM for $k$ parallel multipliers. The architecture is split into sections \texttt{PRNGs} for Trivium modules, \texttt{Error Sampling} for the Gaussian sampler, \texttt{Arithmetic} for the LWE multiplier, and \texttt{Outputs} for the shared-secret and ciphertexts.}\label{arch}
\end{figure}

The LWE matrix multiplication component incurs a large computational overhead. Because of this, it is an ideal target for optimizations, and for our optimizations we heavily rely on parallelisations. Firstly we describe the basic LWE multiplier, that includes just one multiplication component. Then we describe how this core is parallelised, allowing us to significantly improve the throughput.

Figure \ref{arch} shows a high-level overview of the hardware architecture and the following descriptions will link to the design overview. The \texttt{Arithmetic} part of the The LEE core is essentially made by vector-matrix multiplication (that is, $\mathbf{S}[\text{row}]\times \mathbf{A}$), addition of a \texttt{Gaussian} error value (that is, $\mathbf{E}[\text{row},\text{col}]$), and, when needed, an addition of the \texttt{Encoding} of message data. Since the matrix $\mathbf{S}$ consists of a large number of column entries (either 640 or 976) but only 8 row entries (for both parameter sets), we decided to implement a vector-matrix multiplier, instead of (a larger) matrix-matrix one. By doing this, we can reuse the same hardware architecture for each row of $\mathbf{S}$, saving significant hardware resources. Each run of the row-column MAC operation exploits a DSP slice on the FPGA, which fits within the 48-bit MAC size of the FPGA. The DSP slice is ideal for these operations, but it also ensures constant computational run-time, since each multiplication requires one clock cycle. Once each row-column MAC operation is completed, an error value is added from the CDT sampler. These ciphertext values are consistently added into an instantiation of SHAKE, which is required to calculate the shared secret, as well as being output as the ciphertext. This process is pipe-lined to ensure high throughput and constant run-time. A high-level overview of the whole architecture is shown in Figure \ref{arch}.

@@

\begin{figure}[htbp]\centering
    %\advance\leftskip-1cm
\includegraphics[scale=0.65]{figures/parallel.pdf}
\caption{Parallelising matrix multiplication, for $\mathbf{S}\times\mathbf{A}$, used within LWE computations for an example of $k=4$ parallel multiplications, using $k=4$ DSPs on the FPGA.}\label{parallel}
\end{figure}

>>>>

As described in Section \ref{sec:related}, FrodoKEM requires heavy use of PRNG / seed expanding. In the algorithm specifications, it is suggested to use either SHAKE or AES. In particular, the most computationally intensive operations, such as Line \ref{decapslwe} of Algorithm \ref{alg:decaps}, require 410k or 953k 16-bit pseudo-random values, depending on the parameter set used. In order for PRNG not to be the bottle-neck it needs to achieve a very high throughput (ideally with relatively low area consumption) typically in the range of 16 bits per clock cycle. In a previous hardware design, proposed by Howe et al. \cite{howe2018standard}, high throughput for the PRNG was achieved by pre-calculating randomness and storing it in BRAM. Random data newly calculated was then written into the memory, overwriting the random data previously stored. This is an efficient approach, however a more efficient PRNG that would not require BRAM usage, potentially increasing the operating frequency of the design, and thus improve its throughput. Moreover, parallelisations was not possible for this design, as this would require a faster SHAKE design, increasing the area consumption by 3-8x \cite{bertoni2012keccak}. The area consumption of SHAKE (or AES) was an issue with the previous design. For example, cSHAKE used within FrodoKEM-640 Encaps occupies 42\% of the overall hardware resources~\cite{howe2018standard}.

@@

In order to fully explore the potential of FrodoKEM in hardware, we propose several architectures characterized by different design goals (in terms of throughput). We use the proposed architecture to implement key generation, encapsulation, and decapsulation, on two sets of parameters proposed in the specifications: FrodoKEM-640 and FrodoKEM-976. Our designs use 1x, 4x, 8x, and 16x parallel multiplications during the most computationally intensive parts in FrodoKEM. These operations are the LWE matrix multiplications of the form:
%The following is the LWE calculation of the type: 
\begin{equation}
\mathbf{B} = \mathbf{S} \mathbf{A} + \mathbf{E}, \label{lwe}
\end{equation}
 required in key generation, encapsulation, and decapsulation. In the previous hardware implementations of FrodoKEM, the operations of the type of Equation \ref{lwe} took approximately 97.5\% of the overall run-time of the designs \cite{howe2018standard}. As in the literature, we exploit DSP slices on the FPGA for the multiply-and-accumulate (MAC) operations required for matrix multiplication. Hence, each parallel multiplication of the proposed designs requires its own DSP slice. The LWE matrix multiplication component incurs a large computational overhead. Because of this, it is an ideal target for optimizations, and for our optimizations we heavily rely on parallelisations. Firstly we describe the basic LWE multiplier, that includes just one multiplication component. Then we describe how this core is parallelised, allowing us to significantly improve the throughput.

\begin{figure}[htbp]\centering
    %\advance\leftskip-1cm
\includegraphics[scale=0.85]{figures/arch_new.pdf}
\caption{A high-level overview of the proposed hardware designs for FrodoKEM for $k$ parallel multipliers. The architecture is split into sections `PRNGs' for Trivium modules, `Error Sampling' for the Gaussian sampler, `Arithmetic' for the LWE multiplier, and `Outputs' for the shared-secret and ciphertexts.}\label{arch}
\end{figure}

Figure \ref{arch} shows a high-level overview of the hardware architecture and the following descriptions will link to the design overview. The \texttt{Arithmetic} part of the LWE core is essentially made by vector-matrix multiplication (that is, $\mathbf{S}[\text{row}]\times \mathbf{A}$), addition of a \texttt{Gaussian} error value (that is, $\mathbf{E}[\text{row},\text{col}]$), and, when needed, an addition of the \texttt{Encoding} of message data. Since the matrix $\mathbf{S}$ consists of a large number of column entries (either 640 or 976) but only 8 row entries (for both parameter sets), we decided to implement a vector-matrix multiplier, instead of (a larger) matrix-matrix one. By doing this, we can reuse the same hardware architecture for each row of $\mathbf{S}$, saving significant hardware resources. Each run of the row-column MAC operation exploits a DSP slice on the FPGA, which fits within the 48-bit MAC size of the FPGA. The DSP slice is ideal for these operations, but it also ensures constant computational run-time, since each multiplication requires one clock cycle. Once each row-column MAC operation is completed, an error value is added from the CDT sampler. These outputted ciphertext values are also consistently added into an instantiation of SHAKE, which is required to calculate the shared secret. This process is pipe-lined to ensure high throughput and constant run-time.

@@

\begin{figure}[htbp]\centering
    %\advance\leftskip-1cm
\includegraphics[scale=0.52]{figures/parallel.pdf}
\caption{Parallelising matrix multiplication, for $\mathbf{S}\times\mathbf{A}$, used within LWE computations for an example of $k=4$ parallel multiplications, using $k=4$ DSPs on the FPGA.}\vspace{-0.35cm}
\label{parallel}
\end{figure}
